Clustering via Generalized Energy Statistics

### 2018-02-13 energy_clustering/icml18/energy_clustering.tex
icml18 paper

% Use the following line for the initial blind version submitted for review:
%\usepackage{icml2018}

% If accepted, instead use the following line for the camera-ready submission:
\usepackage[accepted]{icml2018}

@@

\begin{abstract}
Energy statistics introduces the notion of potential
energy between probability distributions, in  
close analogy to Newton's gravitational
potential in physics. In this paper, we propose a principled approach to 
clustering based on energy statistics theory.
Our mathematical formulation establishes connection to kernel methods,
leading to a quadratically constrained quadratic program in the associated
feature space.
To obtain local solutions of such an NP-hard optimization problem,
we introduce an iterative algorithm based
on Hartigan's method. This algorithm has the same computational 
cost as kernel $k$-means but offers
several advantages. 
We provide carefully designed numerical experiments illustrating
that the proposed method is more flexible and outperforms
kernel $k$-means, spectral clustering,
standard $k$-means and Gaussian mixture models in a variety of settings,
specially in high dimensions. We also employ the method 
in a real data application describing shapes of neural synapses.
\end{abstract}

@@

\emph{Experimental Setup.} 
The following setup holds unless specified
otherwise. 
We consider
$\mathcal{E}^H$-clustering, kernel $k$-means and spectral clustering
with \mbox{generating} kernels to the following semimetrics:
\begin{subequations}
\label{eq:test_metrics}
\begin{align}
\rho_{\alpha}(x,y) &= \| x-y \|^{\alpha}, \\
\widetilde{\rho}_{\sigma}(x,y) &= 2 - 2 e^{-\tfrac{\|x-y\|}{2 \sigma}},  \\
\widehat{\rho}_{\sigma}(x,y) &= 2 - 2 e^{-\tfrac{\|x-y\|^2}{2 \sigma^2}}.  
\end{align}
\end{subequations}
These are generated by formula \eqref{eq:kernel_semimetric}
and we fix $x_0=0$.
The standard $\rho_1$ will always be present in the experiments 
as a reference, being the implied choice unless explicitly mentioned.
For $k$-means, GMM and spectral clustering we use the robust 
implementations of \emph{scikit-learn} library \citep{scikit-learn}, where  
$k$-means is initialized with 
$k$-means++ \citep{Vassilvitskii}, 
and GMM with the output of $k$-means, making it more robust and preventing
it from breaking in high dimensions. 
Spectral clustering is based on \citet{Malik}. 
We implemented kernel $k$-means
as described in \citet{Dhillon2,Dhillon}, and $\mathcal{E}^H$-clustering
as described in Algorithm~\ref{algo}. Both
will also be initialized with $k$-means++.
We run each algorithm $5$ times with different initializations, picking
the result with best objective function value. 
We evaluate clustering quality by
the \emph{accuracy}, which is the fraction of correctly clustered points
based on the ground truth. More precisely, let $Z$ be the true label
matrix and $\hat{Z}$ the estimated label matrix produced by 
an algorithm. The accuracy is then defined as
\begin{equation}
\label{eq:accuracy}
\textnormal{accuracy}(\hat{Z}) \equiv \max_\pi \dfrac{1}{n} \sum_{i=1}^n
\sum_{j=1}^k \hat{Z}_{i \pi(j)} Z_{ij},
\end{equation}
where $\pi$ is a permutation of the $k$ cluster groups. The accuracy
is always between $[0,1]$, where $1$ correspond to all points correctly
clustered, and $0$ to all points wrongly clustered.
For each setting we show 
the average accuracy over $100$ Monte
Carlo trials, with error bars being standard error.

@@

In Fig.~\ref{fig:other} we have examples of
complex two dimensional datasets. 
We apply $\mathcal{E}^H$-clustering  with different kernel choices,
and also spectral clustering with the best kernel choice, besides 
$k$-means and GMM. Here we perform only $10$ Monte Carlo runs.
For parallel cigars we initialize
all algorithms with $k$-means++, and the concentric circles example
we initialize at random.
The results are shown in Table~\ref{table:other}.
$\mathcal{E}^H$-clustering has superior performance
in every example, in particular better than the
spectral clustering.
For parallel cigars the semimetrics $\rho_1$ and $\rho_{1/2}$ still provide
accurate results, however, for
the concentric circles the kernel choice is more sensitive.

@@

\begin{figure*}
\includegraphics[width=0.18\textwidth]{eigen_gap.pdf}
\includegraphics[width=0.2\textwidth]{heat_means_k6_energy.pdf}
\includegraphics[width=0.2\textwidth]{synapse_cluster_2d_k6_energy.pdf}
\includegraphics[width=0.2\textwidth]{heat_means_k6_gauss.pdf}
\includegraphics[width=0.2\textwidth]{synapse_cluster_2d_k6_gauss.pdf}
\vspace{-0.8cm}
\caption{\label{fig:synapse}
Clustering synapse dataset.
Plots in order from \emph{left} to \emph{right}.
\emph{First:} Eigenvalues of random walk Laplacian. We choose
$k=6$ since its the first time we see a peak. We use energy clustering
with the two semimetrics $\rho_1$ and $\widehat{\rho}_2$; see
\eqref{eq:test_metrics}.
\emph{Second:} Heat map of the cluster means versus features
using energy clustering with standard
metric $\rho_1$ from energy statistics.
\emph{Third:} Clustered points projected into the two principal components,
using $\rho_1$.
\emph{Fourth:} Heat map of the cluster means versus features
using energy clustering with Gaussian semimetric 
$\widehat{\rho}_2$.
\emph{Fifth:} Clustered points projected into the two principal components,
using $\widehat{\rho}_2$.
}
\end{figure*}

We now illustrate an application to real data of 
$\mathcal{E}^H$-clustering algorithm based on Hartigan's method.
This dataset describes the shapes of neural synapses and was collected
by expert neuroscientists. It consists of $1025$ data points
with $11$ features. Each point
corresponds to $(x,y,z)$ location in the brain. We normalize this
dataset beforehand.

We apply $\mathcal{E}^H$-clustering using the two
semimetrics $\rho_1$ and $\widehat{\rho}_2$; see \eqref{eq:test_metrics}.
We compute the difference between eigenvalues of the random walk Laplacian
obtained from the kernel matrix $G$ \eqref{eq:kernel_matrix}.
The choice of $k$ corresponds to the first time we see a meaningfull
peak in the plot, which occurs at $k=6$; first plot in
Fig.~\ref{fig:synapse}.
This procedure to find $k$ is commonn in the literature \cite{vonLuxburg2007}.
Having found $k$ we now cluster the data.
First we use the semimetric $\rho_1$, which
is the standard Euclidean choice in energy statistics.
In the second and third plots of Fig.~\ref{fig:synapse} we show a heat
map of the cluster means versus features and the clustered data points
projected into the 2 principal components, respectivelly.
The fourth and fifth plots in Fig.~\ref{fig:synapse} show the same
experiment but using the semimetric $\widehat{\rho}_2$ corresponding
to a Gaussian kernel.

@@

We proposed clustering from the perspective of generalized energy
statistics, valid for arbitrary spaces of negative type.
Our mathematical formulation yields
a QCQP in the associated kernel Hilbert space; 
see  Proposition~\ref{th:qcqp2}.
We demonstrated that such QCQP
is equivalent to kernel $k$-means optimization problem, 
once the kernel is fixed; see
Proposition~\ref{th:kernel_kmeans}. Energy statistics, however, fixes
a family of standard kernels in Euclidean space, and
more general kernels 
on spaces of negative type can also be obtained.

We proposed the iterative $\mathcal{E}^H$-clustering 
(Algorithm~\ref{algo})
which is 
a kernelized version of Hartigan's method. 
It was compared to kernel $k$-means algorithm,
based on Lloyd's heuristic.
Both have the same time complexity, however, numerical and theoretical
results provide compelling evidence that $\mathcal{E}^H$-clustering
is more robust with a superior performance, specially in high
dimensions. 
Furthermore, $\mathcal{E}^H$-clustering with standard kernels from energy
statistics outperformed $k$-means and GMM
on several settings, even on Gaussian ones.
In some cases
$\mathcal{E}^H$-clustering also surpassed spectral clustering, and in
others performed similarly but never worse. 
Note that computing eigenvectors
of large matrices can be unfeasible and iterative methods are preferred.
We provided an application of $\mathcal{E}^H$-clustering to a real dataset
describing shapes of neural synapses.
Finally, we remark that kernel methods can benefit from sparsity and
fixed-rank approximations of the Gram matrix and there is plenty
of room to make $\mathcal{E}^H$-clustering more scalable.

\subsection*{Acknowledgements}
We would like to thank Carey Priebe 
for discussions.
We would like to acknowledge the support of the Transformative
Research Award (NIH \#R01NS092474) and  the Defense Advanced Research Projects
Agency’s (DARPA) SIMPLEX program through SPAWAR contract N66001-15-C-4041.

>>>>

% Use the following line for the initial blind version submitted for review:
\usepackage{icml2018}

% If accepted, instead use the following line for the camera-ready submission:
%\usepackage[accepted]{icml2018}

@@

\begin{abstract}
Energy statistics introduces the notion of potential
energy between probability distributions, in  
close analogy to Newton's gravitational
potential in physics. In this paper, we propose a principled approach to 
clustering based on energy statistics theory.
Our mathematical formulation establishes connection to kernel methods,
leading to a quadratically constrained quadratic program in the associated
feature space.
To obtain local solutions of such an NP-hard optimization problem,
we introduce an iterative algorithm based
on Hartigan's method. This algorithm has the same computational 
cost as kernel $k$-means but offers
several advantages. 
We provide carefully designed numerical experiments illustrating
that the proposed method is more flexible and outperforms
kernel $k$-means, spectral clustering,
standard $k$-means and Gaussian mixture models in a variety of settings,
specially in high dimensions. We employ the method 
to an important real dataset describing protein expressions
of neural synapses.
\end{abstract}

@@

\emph{Experimental Setup.} 
The following setup holds unless specified
otherwise. 
We consider
$\mathcal{E}^H$-clustering, kernel $k$-means and spectral clustering
with \mbox{generating} kernels to the following metrics:
\begin{subequations}
\label{eq:test_metrics}
\begin{align}
\rho_{\alpha}(x,y) &= \| x-y \|^{\alpha}, \\
\widetilde{\rho}_{\sigma}(x,y) &= 2 - 2 e^{-\tfrac{\|x-y\|}{2 \sigma}},  \\
\widehat{\rho}_{\sigma}(x,y) &= 2 - 2 e^{-\tfrac{\|x-y\|^2}{2 \sigma^2}}.  
\end{align}
\end{subequations}
These are generated by formula \eqref{eq:kernel_semimetric}
and we fix $x_0=0$.
The standard $\rho_1$ will always be present in the experiments 
as a reference, being the implied choice unless explicitly mentioned.
For $k$-means, GMM and spectral clustering we use the robust 
implementations of \emph{scikit-learn} library \citep{scikit-learn}, where  
$k$-means is initialized with 
$k$-means++ \citep{Vassilvitskii}, 
and GMM with the output of $k$-means, making it more robust and preventing
it from breaking in high dimensions. 
Spectral clustering is based on \citet{Malik}. 
We implemented kernel $k$-means
as described in \citet{Dhillon2,Dhillon}, and $\mathcal{E}^H$-clustering
as described in Algorithm~\ref{algo}. Both
will also be initialized with $k$-means++.
We run each algorithm $5$ times with different initializations, picking
the result with best objective function value. 
We evaluate clustering quality by
the \emph{accuracy}, which is the fraction of correctly clustered points
based on the ground truth. More precisely, let $Z$ be the true label
matrix and $\hat{Z}$ the estimated label matrix produced by 
an algorithm. The accuracy is then defined as
\begin{equation}
\label{eq:accuracy}
\textnormal{accuracy}(\hat{Z}) \equiv \max_\pi \dfrac{1}{n} \sum_{i=1}^n
\sum_{j=1}^k \hat{Z}_{i \pi(j)} Z_{ij},
\end{equation}
where $\pi$ is a permutation of the $k$ cluster groups. The accuracy
is always between $[0,1]$, where $1$ correspond to all points correctly
clustered, and $0$ to all points wrongly clustered.
For each setting we show 
the average accuracy over $100$ Monte
Carlo trials, with error bars being standard error.

@@

In Fig.~\ref{fig:other} we have examples of
complex two dimensional datasets. 
We apply $\mathcal{E}^H$-clustering  with different kernel choices,
and also spectral clustering with the best kernel choice, besides 
$k$-means and GMM. Here we perform only $10$ Monte Carlo runs.
For parallel cigars we initialize
all algorithms with $k$-means++, and the concentric circles example
we initialize at random.
The results are shown in Table~\ref{table:other}.
$\mathcal{E}^H$-clustering has superior performance
in every example, in particular better than the
spectral clustering.
For parallel cigars the metrics $\rho_1$ and $\rho_{1/2}$ still provide
accurate results, however, for
the concentric circles the kernel choice is more sensitive.

@@

\begin{figure*}
\includegraphics[width=0.18\textwidth]{eigen_gap.pdf}
\includegraphics[width=0.2\textwidth]{heat_means_k6_energy.pdf}
\includegraphics[width=0.2\textwidth]{synapse_cluster_2d_k6_energy.pdf}
\includegraphics[width=0.2\textwidth]{heat_means_k6_gauss.pdf}
\includegraphics[width=0.2\textwidth]{synapse_cluster_2d_k6_gauss.pdf}
\vspace{-0.8cm}
\caption{\label{fig:synapse}
Clustering synapse dataset.
Plots in order from \emph{left} to \emph{right}.
\emph{First:} Eigenvalues of random walk Laplacian. We choose
$k=6$ since its the first time we see a peak. We use energy clustering
with the two metrics $\rho_1$ and $\widehat{\rho}_2$; see
\eqref{eq:test_metrics}.
\emph{Second:} Heat map of the cluster means versus features
using energy clustering with standard
metric $\rho_1$ from energy statistics.
\emph{Third:} Clustered points projected into the two principal components,
using $\rho_1$.
\emph{Fourth:} Heat map of the cluster means versus features
using energy clustering with Gaussian metric 
$\widehat{\rho}_2$.
\emph{Fifth:} Clustered points projected into the two principal components,
using $\widehat{\rho}_2$.
}
\end{figure*}

The considered dataset 
describes protein expression of neural synapses (chemical 
connections between neurons), being
crucial for understanding the diversity of synapses. This is
part of a large NIH funded consortium within the BRAIN initiative;
we refer to \citet{Collman2015} for details.
We have $1025$ data points
with $11$ features. Each point
corresponds to a $(x,y,z)$ location in the brain. We normalize this
dataset before applying $\mathcal{E}^H$-clustering.

We use the two
metrics $\rho_1$ and $\widehat{\rho}_2$; see \eqref{eq:test_metrics}.
We compute the difference between eigenvalues of the random walk Laplacian
obtained from the kernel matrix $G$ \eqref{eq:kernel_matrix}.
The choice of $k$ corresponds to the first time we see a meaningful
peak in the plot, which occurs at $k=6$; first plot in
Fig.~\ref{fig:synapse}.
This procedure to find $k$ is common in the literature \cite{vonLuxburg2007}.
Having found $k$ we now cluster the data.
First we use the metric $\rho_1$, which
is the standard Euclidean choice in energy statistics.
In the second and third plots of Fig.~\ref{fig:synapse} we show a heat
map of the cluster means versus features and the clustered data points
projected into the 2 principal components, respectively.
The fourth and fifth plots in Fig.~\ref{fig:synapse} show the same
experiment but using the metric $\widehat{\rho}_2$ corresponding
to a Gaussian kernel.

@@

We proposed clustering from the perspective of generalized energy
statistics, valid for arbitrary spaces of negative type.
Our mathematical formulation yields
a QCQP in the associated RKHS;
Proposition~\ref{th:qcqp2}.
We showed that such QCQP
is equivalent to kernel $k$-means optimization problem, 
once the kernel is fixed; 
Proposition~\ref{th:kernel_kmeans}. However, energy statistics fixes
a family of standard kernels in Euclidean space, and
more general kernels 
on spaces of negative type as well.

We proposed the iterative $\mathcal{E}^H$-clustering 
(Algorithm~\ref{algo})
which is 
a kernelized version of Hartigan's method. 
It was compared to kernel $k$-means algorithm,
based on Lloyd's heuristic.
Both have the same complexity, however, numerical and theoretical
results provide compelling evidence that $\mathcal{E}^H$-clustering
is more robust with superior performance, specially in high
dimensions. 
Moreover, $\mathcal{E}^H$-clustering with standard kernels from energy
statistics outperformed $k$-means and GMM
on several settings, even on Gaussian ones.
In some cases
$\mathcal{E}^H$-clustering also surpassed spectral clustering, and in
others performed similarly but never worse. 
Note that computing eigenvectors
of large matrices can be unfeasible and iterative methods are preferred.
We provided an application of $\mathcal{E}^H$-clustering to an
important real dataset
describing protein expression of neural synapses.
Finally, we remark that kernel methods can benefit from sparsity and
fixed-rank approximations of the Gram matrix and there is plenty
of room to make $\mathcal{E}^H$-clustering more scalable.

%\subsection*{Acknowledgements}
%We would like to thank Carey Priebe 
%for discussions.
%We would like to acknowledge the support of the Transformative
%Research Award (NIH \#R01NS092474) and  the Defense Advanced Research Projects
%Agency’s (DARPA) SIMPLEX program through SPAWAR contract N66001-15-C-4041.
